### 2023-09-27 submission/manuscript.tex
Submit to journal
Respond to reviewers' comments

In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
not, want not: why rarefying microbiome data is inadmissible''. The
claims of their study have significantly altered how microbiome
researchers control for the unavoidable uneven sequencing depths that
are inherent in modern 16S rRNA gene sequencing. Confusion over the
distinction between the definitions of rarefying and rarefaction
continues to cloud the interpretation of their results. More
importantly, the authors made a variety of choices in designing and
analyzing their simulations that were potentially problematic. I
identified 11 factors that could have compromised the results of the
original study. I reproduced the original simulation results and
assessed the impact of those 11 factors on the underlying conclusion
that rarefying data is inadmissible. Throughout, the design of the
original study made choices that caused rarefying and rarefaction to
appear to perform worse than they truly did. Most important were the
approaches used to assess ecological distances, the removal of samples
with low sequencing depth, and not accounting for conditions where
sequencing effort is confounded with treatment group. Although the
original study criticized rarefying for the arbitrary removal of valid
data, repeatedly rarefying data many times (i.e., rarefaction)
incorporates all the data. In contrast, it is the removal of rare taxa
that would appear to remove valid data. Overall, I show that rarefaction
is the most robust approach to control for uneven sequencing effort when
considered across a variety of alpha and beta diversity metrics.

@@

Over the past 10 years, the best method for normalizing the sequencing
depth of samples characterized by 16S rRNA gene sequencing has been
contentious. An often cited paper by McMurdie and Holmes forcefully
argued that rarefying the number of sequence counts was ``inadmissible''
and should not be employed. However, I identified a number of problems
with the design of their simulations and analysis that could compromise
their results. In fact, when I reproduced and expanded upon their
analysis, it was clear that rarefaction was actually the most robust
approach for controlling for uneven sequencing effort across samples.
Rarefaction limits the rate of falsely detecting and rejecting
differences between treatment groups. Far from being ``inadmissible'',
rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences (1--3). These barcodes allow
researchers to pool PCR products and then deconvolute the resulting
sequence data based on the barcode sequences. Despite researchers' best
efforts to generate equimolar pools of PCR products, it is common to
observe more than 10-fold variation in the number of sequences per
sample (4). Researchers desire strategies to minimize uneven sequencing
depth and thus need methods to control for this unevenness in their
analyses. Of course, uneven sequencing is not unique to microbiome
research and is a challenge faced by all community ecologists (5, 6).
Common approaches to control for the effects of uneven sequencing depths
include use of proportional abundance (i.e., relative abundance),
scaling of counts, parameter estimation, and rarefaction (7--23).

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (24). Their paper has had a significant impact on
the approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
that showed how rarefaction is beneficial in some cases (25); however,
the proponents of WNWN appear to be holding sway over the microbiome
community. I have received correspondence from researchers over the past
10 years asking how to address critiques from reviewers who criticize my
correspondents' analysis for using rarefaction (e.g., see
\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
have also received similar comments from reviewers regarding my own
work. For example, I received the critique for a preprint that I posted
in 2020 that analyzed the practice of removing rare taxa from microbiome
analyses {[}see comments linked to (4);
\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
In the process of responding to these reviewers' comments and preparing
a manuscript investigating rarefaction and other approaches to control
for uneven sequencing effort, I grew to appreciate that there is
significant confusion in the field over what is meant by ``rarefying''
and ``rarefaction''.

McMurdie and Holmes discussed and defined ``rarefying'' in WNWN
(citation numbers updated from the original to those of the current
study):

\begin{quote}
``Instead, microbiome analysis workflows often begin with an ad hoc
library size normalization by random subsampling without replacement, or
so-called rarefying (13, 26, 27). There is confusion in the literature
regarding terminology, and sometimes this normalization approach is
conflated with a non-parametric resampling technique --- called
rarefaction (5), or individual-based taxon re-sampling curves (28) ---
that can be justified for coverage analysis or species richness
estimation in some settings (28), though in other settings it can
perform worse than parametric methods (29). Here we emphasize the
distinction between taxon re-sampling curves and normalization by
strictly adhering to the terms rarefying or rarefied counts when
referring to the normalization procedure, respecting the original
definition for rarefaction. Rarefying is most often defined by the
following steps (13).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. This has also
  been called the rarefaction level (26), though we will not use the
  term here.
\item
  Discard libraries (microbiome samples) that have fewer reads than
  N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\end{enumerate}

Often N\textsubscript{L,m} is chosen to be equal to the size of the
smallest library that is not considered defective, and the process of
identifying defective samples comes with a risk of subjectivity and
bias. In many cases researchers have also failed to repeat the random
subsampling step (3) or record the pseudorandom number generation
seed/process --- both of which are essential for reproducibility.''
\end{quote}

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
in this quote and throughout their manuscript. The authors were correct
to state that the distinction between ``rarefying'' and ``rarefaction''
is confusing and leads to their conflation. Adding to the confusion is
that the papers cited in the first sentence of this quote (i.e., (13,
26, 27)) either do not use the words ``rarefy'' or ``rarefying'' or use
them interchangeably with ``rarefaction''. For example, Hughes and
Hellman did not use ``rarefy'' and use ``rarefaction'' in the
traditional sense with multiple subsamplings (13). Meanwhile, the
QIIME-based literature appears to use ``rarefy'' and ``rarefaction''
interchangeably to mean only a single subsampling (26, 27). Subsequent
researchers have continued to conflate the terms when citing WNWN
(\textbf{Supplemental Text}). An exemplar of the confusion is the
creation of a technique that uses ``repeatedly rarefying'' as an
approach distinct from rarefaction when they were re-proposing
traditional rarefaction (30).

As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize
the distinction between rarefying and rarefaction. However, because they
seem to have coined a new definition for rarefying, they added to the
confusion by using the generally used verb form of rarefaction. Further
confusion comes from the author's admonition in the final sentence of
the above quoted text that some researchers have failed to repeat the
subsampling step. Traditionally, repeating the subsampling step a large
number of times and averaging the result is rarefaction (12, 13, 28). In
other words, rarefying or subsampling is rarefaction, but with a single
randomization. To minimize confusion, I will use ``subsampling'' in
place of ``rarefying'' through the remainder of this study.

For clarity, I will use the following definition of rarefaction:

@@

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sequenced to different depths (5, 6).
With this more general approach to the definition of rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, consideration is needed to synthesize the results of
these tests across a large number of replications.

@@

For the WNWN cluster analysis, OTU abundance distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPatterns dataset (L129) (2). To generate
the fecal and ocean distributions, the authors included any operational
taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
samples the OTUs were observed in followed by their total abundance
across all 7 samples (L139). From this sorted list they obtained the
identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
they selected the data for the corresponding 2,000 OTUs and pooled the
OTU abundances of the fecal and ocean samples separately to create two
OTU abundance distributions (L144, L159-160, L197-198). Next, the fecal
and ocean distributions were mixed in 8 different fractions to generate
two community types that differed by varying effect sizes (i.e., 1.00,
1.15, 1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect
size of 1.00 generated a null model with no difference between the
treatment groups. To simulate the variation in sequencing depth across
the 80 samples, they normalized the number of sequences from each of the
26 samples in the GlobalPatterns dataset so that the median number of
sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80
sequencing depths. From each community type, they simulated 40 samples
by sampling to the desired number of sequences (L73, L230-233 and
L326-327). Each simulation condition was repeated 5 times (L85).
Finally, they removed rare and low prevalence OTUs in two steps. First,
they removed any OTUs whose total abundance was less than 3 across all
80 samples and that did not appear in at least 3 samples (L368-386).
Second, they removed any OTUs that did not have more than 1 sequence in
more than 5\% of the 80 samples (i.e., 4 samples) and that did not have
a total abundance across the 80 samples greater than one half of the
number of samples in each community type (i.e., 20) (L523-538, L551).
Simulation A consisted of 160 simulations (8 effect sizes x 4 median
sequencing depths x 5 replicates = 160 simulations).

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean,
Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
and Weighted UniFrac distances (L760-767). Subsampled data were used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
distances, and top-Mean Squared Difference (L769-775, L793-798,
L801-806). DESeq Variance Stabilization normalized data were used to
calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
used to calculate top-Mean Squared Difference distances (L801-806). WNWN
calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
package (34). They calculated Poisson distances using the
\texttt{PoiClaClu} R package (35). The top-mean squared difference was
calculated using the \texttt{edgeR} R package (19). The resulting
distance matrices were used to cluster the 80 samples into one of two
clusters using partitioning around the medoid (PAM), K-means clustering,
and hierarchical clustering (L865-879). Although data for all three
methods were presented in Protocol S1, only the PAM data were presented
in the main manuscript. The accuracy of the clustering assignments was
quantified as the fraction of the 80 samples that were assigned to the
correct cluster (L887-908). Since some of the subsampling conditions
removed samples below a minimum sequencing depth threshold, those
samples were counted as mis-clustered samples yielding minimum
accuracies below 50\%.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, potentially
  increasing the sensitivity of results to the choice of the random
  number generator seed
\item
  The average sizes of the libraries were small by modern standards,
  which may limit the generalizability of the results
\item
  DESeq-based Variance Stabilization generates negative values and was
  used with distance calculation methods that are sensitive to negative
  values, which likely led to nonsense distances and clusters
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering, although close inspection
  suggests that K-means may have been superior to PAM for some
  conditions
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses, which may limit the generalizability of
  the results
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which may have skewed the shape of the community
  distributions
\item
  There was no accounting for differences in performance when library
  sizes are confounded with treatment group
\item
  Clustering accuracy was used rather than the more direct and
  frequently applied comparisons of beta diversity using permutation
  tests
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics, which is the traditional application of
  rarefaction
\end{enumerate}

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of WNWN using the
authors' code. I created a Conda environment that used the R version and
package versions that were as close as possible to those used in WNWN.
It was necessary to patch the WNWN's R markdown file to render the
document to be compatible with the overall workflow of this study. I was
able to generate figures similar to those presented as WNWN's Figures 4
and 5; my results are shown in this study as Figures 1 and 2,
respectively. The differences in results are likely due to differences
in software versions and operating systems. It is also worth noting that
the published versions of the two figures differ from those included in
Protocol S1 within the rendered HTML file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R markdown file (i.e., Protocol S1 labels the figures
as Figures 3 and 4 corresponding to the published Figures 4 and 5).
Regardless of the differences, my results were qualitatively similar to
that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reported the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 10 and 11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or interquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R markdown document. Again, the observed
results were qualitatively similar to those generated using the single R
markdown file (Figures 3 and 4). The increased number of replications
resulted in smoother lines and allowed me to present empirical 95\%
confidence intervals. For all analyses in the remainder of this paper, I
used 100 randomized replicates per condition.

\hypertarget{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}{%
\subsubsection{2. The average sizes of the libraries were small by
modern standards, which may limit the generalizability of the
results}\label{the-average-sizes-of-the-libraries-were-small-by-modern-standards-which-may-limit-the-generalizability-of-the-results}}

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
an additional median depth of sequencing of 50,000 sequences with WNWN's
four median sequencing depths (i.e., 1,000, 2,000, 5,000, 10,000).
Additional sequencing coverage would be expected to result in more
robust distance values since there would be more information represented
in the data. Indeed, the added sequencing depth showed higher accuracy
values at lower effect sizes for the combinations of normalization
methods and distance calculations (Figure 3). Increased sequencing
coverage also resulted in improved clustering accuracy for lower effect
sizes when the library size minimum quantile was decreased (Figure 4). I
will revisit the choice of the library size minimum quantile below.

@@

A more robust analysis would have used rarefaction since it would have
averaged across a large number of random subsamplings (e.g., 100 or
1,000). By using a large number of subsamplings the likelihood of
incorporating all of the OTUs would have increased. Rather than being
guilty of ``omission of available valid data'' as claimed in WNWN, with
a sufficient number of subsamplings, traditional rarefaction uses all of
the available data. To fairly compare subsampling, as employed in WNWN,
and rarefaction, I removed the 15\% of samples with the lowest number of
sequences and compared the clustering accuracies from a single
subsampling to rarefaction with 100 randomizations. This analysis
revealed two benefits of rarefaction. First, the median distances
generated by rarefaction was always at least as large as those from a
single subsample (Figure 6). The difference was most pronounced for
smaller average library sizes and at smaller effect sizes. The
Unweighted UniFrac distances were most impacted by the use of
rarefaction over subsampling. Second, the interquartile ranges in
clustering accuracy by rarefaction were generally smaller than those by
subsampling and showed similar trends to the difference in the median
distances (Figure 6). Because rarefaction incorporates more of the data
and generally performed better than subsampling, the remainder of this
analysis will report results using rarefaction rather than by
subsampling, except when noted.

@@

Using the \texttt{nr-s1} Conda environment, I measured the richness and
Shannon diversity using the normalized OTU counts. Richness was measured
by counting the number of OTUs for each sample across the simulations
and the Shannon diversity using their relative abundance using the
commonly used formula (31). For the rarefaction data, each of the 80
samples in the 100 subsampled datasets were used to calculate the
richness and diversity. The mean across the 100 subsamplings was used as
the value for the rarefaction data across each of the 80 samples. To
assess differences between the two treatment groups, the richness and
diversity values were compared using the two-sample Wilcoxon
non-parametric test with the \texttt{wilcox.test} from the
\texttt{stats} base R package. Similar to the analysis using
\texttt{adonis2}, the fraction of significant tests was the fraction of
the 100 replicate simulations that had a p-value less than or equal to
0.05. The Type I error rate for a condition was the fraction of
significant tests when the effect size was 1.00. The power was the
fraction of significant tests at other effect sizes.

@@

\includegraphics{figure_02.png}

\textbf{Figure 2. Re-running the R markdown provided in Protocol S1 of
WNWN qualitatively reproduced Figure 5 from WNWN.}

\includegraphics{figure_03.png}

>>>>

In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
not, want not: why rarefying microbiome data is inadmissible''. The
claims of their study have significantly altered how microbiome
researchers control for the unavoidable uneven sequencing depths that
are inherent in modern 16S rRNA gene sequencing. Confusion over the
distinction between the definitions of rarefying and rarefaction
continues to cloud the interpretation of their results. More
importantly, the authors made a variety of problematic choices when
designing and analyzing their simulations. I identified 11 factors that
could have compromised the results of the original study. I reproduced
the original simulation results and assessed the impact of those factors
on the underlying conclusion that rarefying data is inadmissible.
Throughout, the design of the original study made choices that caused
rarefying and rarefaction to appear to perform worse than they truly
did. Most important were the approaches used to assess ecological
distances, the removal of samples with low sequencing depth, and not
accounting for conditions where sequencing effort is confounded with
treatment group. Although the original study criticized rarefying for
the arbitrary removal of valid data, repeatedly rarefying data many
times (i.e., rarefaction) incorporates all the data. In contrast, it is
the removal of rare taxa that would appear to remove valid data.
Overall, I show that rarefaction is the most robust approach to control
for uneven sequencing effort when considered across a variety of alpha
and beta diversity metrics.

@@

Over the past 10 years, the best method for normalizing the sequencing
depth of samples characterized by 16S rRNA gene sequencing has been
contentious. An often cited paper by McMurdie and Holmes forcefully
argued that rarefying the number of sequence counts was ``inadmissible''
and should not be employed. However, I identified a number of problems
with the design of their simulations and analysis that compromised their
results. In fact, when I reproduced and expanded upon their analysis, it
was clear that rarefaction was actually the most robust approach for
controlling for uneven sequencing effort across samples. Rarefaction
limits the rate of falsely detecting and rejecting differences between
treatment groups. Far from being ``inadmissible'', rarefaction is a
valuable tool for analyzing microbiome sequence data.

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences (1--3). These barcodes allow
researchers to pool PCR products and then deconvolute the resulting
sequence data based on the barcode sequences. Despite researchers' best
efforts to generate equimolar pools of PCR products, it is common to
observe more than 10-fold variation in the number of sequences per
sample (4). Researchers desire strategies to minimize uneven sequencing
depth and thus need methods to control for this unevenness in their
analyses. Of course, uneven sampling effort is not unique to microbiome
research and is a challenge faced by all community ecologists (5, 6).
Common approaches to control for the effects of uneven sequencing depths
include use of proportional abundance (i.e., relative abundance),
scaling of counts, parameter estimation, and rarefaction (7--23).

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (24). The provocative title attempts to express
the idea that if one does not waste resources then they will not be
found wanting the resource when it is needed. Applied to their study, it
implies that rarefying data arbitrarily discards samples and sequence
data in the subsampling step leading to a loss of statistical power.
Their paper has had a significant impact on the approaches that
microbiome researchers use to analyze 16S rRNA gene sequence data.
According to Google Scholar, this paper has been cited more than 2,560
times as of October 2023. There has been a rebuttal of WNWN that showed
how rarefaction is beneficial in some cases (25); however, the
proponents of WNWN appear to be holding sway over the microbiome
community. I have received correspondence from researchers over the past
10 years asking how to address critiques from reviewers who criticize my
correspondents' analysis for using rarefaction. I have also received
similar comments from reviewers regarding my own work. In responding to
such critiques, I have grown to appreciate that there is significant
confusion in the field over what is meant by ``rarefying'' and
``rarefaction''.

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
throughout their manuscript. The authors were correct to state that the
distinction between ``rarefying'' and ``rarefaction'' is confusing and
leads to their conflation. They defined rarefying as taking a subsample
of the same number of reads from each sample (N\textsubscript{L,m})
without replacement and discarding any samples that have fewer than that
number of reads. The subsampling is performed without replacement.
Traditionally, rarefaction involves repeating the subsampling step a
large number of times, calculating a metric, and averaging the metric
across the subsamples (12, 13, 26). In other words, rarefying or
subsampling is rarefaction, but with a single randomization.

Confusion over the terms is demonstrated in the choice of citations that
McMurdie and Holmes used to define rarefying (i.e., (13, 27, 28)). The
cited studies either did not use the words ``rarefy'' or ``rarefying''
or used them interchangeably with rarefy as a verb form of rarefaction.
For example, Hughes and Hellman did not use ``rarefy'' (13). Rather they
used ``rarefaction'' in the traditional sense with multiple
subsamplings. Meanwhile, the QIIME-based literature appears to use
``rarefy'' and ``rarefaction'' interchangeably to mean only a single
subsampling (27, 28). Confusion comes from the WNWN authors' admonition
that ``{[}i{]}n many cases researchers have also failed to repeat the
random subsampling step''. This seems to call on researchers to use
rarefaction rather than the single subsampling step. Subsequent
researchers have continued to conflate the terms when citing WNWN
(\textbf{Supplemental Text}). An exemplar of the confusion is the
creation of a technique that uses ``repeatedly rarefying'' as an
approach distinct from rarefaction when they were in fact re-proposing
traditional rarefaction (29).

To minimize confusion, I will use ``subsampling'' in place of
``rarefying'' through the remainder of this study and I will use the
following definition of rarefaction:

@@

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sequenced to different depths (5, 6).
With this more general approach to the definition of rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, further consideration is needed to synthesize the
results of these tests across a large number of replications.

@@

For the WNWN cluster analysis, OTU abundances and sequencing depths were
obtained from the GobalPatterns dataset (2), which consisted of 26
samples originally obtained from 9 environments including creek (n=3),
human feces (n=4), human tongue (n=2), lake (n=2), mock communities
(n=3), ocean (n=3), sediment (n=3), skin (n=3), and soil (n=3). In the
WNWN analysis, the 7 human fecal and ocean sequence datasets were used
to generate the OTU distributions for the two simulated treatment groups
(L129). To generate the fecal and ocean distributions, the authors
included any operational taxonomic unit (OTU) that appeared in more than
one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were
sorted by how many of the 7 samples the OTUs were observed in followed
by their total abundance across all 7 samples (L139). From this sorted
list they obtained the identifiers of the first 2,000 OTUs (L66).
Returning to the 7 samples they selected the data for the corresponding
2,000 OTUs and pooled the OTU abundances of the fecal and ocean samples
separately to create two OTU abundance distributions (L144, L159-160,
L197-198). Next, the fecal and ocean distributions were mixed in 8
different fractions to generate two community types that differed by
varying effect sizes (i.e., 1.00, 1.15, 1.25, 1.50, 1.75, 2.00, 2.50,
and 3.50; L170-195, L220); an effect size of 1.00 generated a null model
with no difference between the treatment groups. To simulate the
variation in sequencing depth across the 80 samples, they normalized the
number of sequences from each of the 26 samples in the full
GlobalPatterns dataset so that the median number of sequences
(Ñ\textsubscript{L}) across the samples had 1,000, 2,000, 5,000, or
10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80
sequencing depths. From each community type, they simulated 40 samples
by sampling to the desired number of sequences (L73, L230-233 and
L326-327). Each simulation condition was repeated 5 times (L85).
Finally, they removed rare and low prevalence OTUs in two steps. First,
they removed any OTUs whose total abundance was less than 3 across all
80 samples and that did not appear in at least 3 samples (L368-386).
Second, they removed any OTUs that did not have more than 1 sequence in
more than 5\% of the 80 samples (i.e., 4 samples) and that did not have
a total abundance across the 80 samples greater than one half of the
number of samples in each community type (i.e., 20) (L523-538, L551).
Simulation A consisted of 160 simulations (8 effect sizes x 4 median
sequencing depths x 5 replicates = 160 simulations).

@@

Non-normalized data were used to calculate Bray-Curtis, Euclidean,
Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
and Weighted UniFrac distances (L760-767). Subsampled data were used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
distances, and top-Mean Squared Difference (L769-775, L793-798,
L801-806). DESeq Variance Stabilization normalized data were used to
calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
used to calculate top-Mean Squared Difference distances (L801-806). WNWN
calculated Bray-Curtis (30), Euclidean (30), Unweighted UniFrac (31),
and Weighted UniFrac distances (32) using the \texttt{Phyloseq} R
package (33). They calculated Poisson distances using the
\texttt{PoiClaClu} R package (34). The top-Mean Squared Difference was
calculated using the \texttt{edgeR} R package (19). The resulting
distance matrices were used to cluster the 80 samples into one of two
clusters using partitioning around the medoid (PAM), K-means clustering,
and hierarchical clustering (L865-879). Although data for all three
methods were presented in Protocol S1, only the PAM data were presented
in the main manuscript. The accuracy of the clustering assignments was
quantified as the fraction of the 80 samples that were assigned to the
correct cluster (L887-908). Since some of the subsampling conditions
removed samples below a minimum sequencing depth threshold, the removed
samples were counted as mis-clustered samples yielding minimum
accuracies below 50\%.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, potentially
  increasing the sensitivity of results to the choice of the random
  number generator seed
\item
  The average sizes of the libraries do not cover the larger sequencing
  depths frequently found in modern microbiome studies, which may limit
  the generalizability of the results
\item
  DESeq-based Variance Stabilization generates negative values and was
  used with distance calculation methods that are sensitive to negative
  values, which likely led to nonsense distances and clusters
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering, although close inspection
  suggests that K-means may have been superior to PAM for some
  conditions
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses, which may limit the generalizability of
  the results
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which may have skewed the shape of the community
  distributions
\item
  There was no accounting for differences in performance when library
  sizes are confounded with treatment group
\item
  Clustering accuracy was used rather than the more direct and
  frequently applied comparisons of beta diversity using permutation
  tests
\item
  There was no consideration of effects of normalization methods on
  richness, which is the traditional application of rarefaction
\end{enumerate}

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of WNWN using the
authors' code. I created a Conda environment that used the R version and
package versions that were as close as possible to those used in WNWN.
It was necessary to patch the WNWN's R markdown file to render the
document to be compatible with the overall workflow of this study. I was
able to generate figures similar to those presented as WNWN's Figures 4
and 5; my results are shown in this study as Figures S1 and S2,
respectively. The differences in results are likely due to differences
in software versions and operating systems. It is also worth noting that
the published versions of the two figures differ from those included in
Protocol S1 within the rendered HTML file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R markdown file (i.e., Protocol S1 labels the figures
as Figures 3 and 4 corresponding to the published Figures 4 and 5).
Regardless of the differences, my results were qualitatively similar to
that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reported the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeq Variance Stabilization normalized
data). A better approach would have been to use 100 replicates as this
would reduce the dependency of the results on the random number
generator's seed. By increasing the number of replicates it was also
possible to compare the probability of falsely and correctly clustering
samples from the same and different treatment groups together (see
points 10 and 11, below). Because the accuracies were not symmetric
around the mean accuracy values, the median and 95\% confidence
intervals or interquartile range should have been reported. To test the
effect of increasing the number of replicates, I pulled apart the code
in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R markdown document. Again, my results were
qualitatively similar to those generated using the single R markdown
file (Figures S3 and S4). The increased number of replications resulted
in smoother lines and allowed me to present empirical 95\% confidence
intervals. For all analyses in the remainder of this paper, I used 100
randomized replicates per condition.

\hypertarget{the-average-sizes-of-the-libraries-do-not-cover-the-larger-sequencing-depths-frequently-found-in-modern-microbiome-studies-which-may-limit-the-generalizability-of-the-results}{%
\subsubsection{2. The average sizes of the libraries do not cover the
larger sequencing depths frequently found in modern microbiome studies,
which may limit the generalizability of the
results}\label{the-average-sizes-of-the-libraries-do-not-cover-the-larger-sequencing-depths-frequently-found-in-modern-microbiome-studies-which-may-limit-the-generalizability-of-the-results}}

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. Although
sequencing depths at the smaller end of the range used in WNWN are
reported in recent studies, it is increasingly common to find a median
number of sequences larger than 10,000 (see Table 1 of (4)). Therefore,
I included an additional median depth of sequencing of 50,000 sequences
with WNWN's four median sequencing depths (i.e., 1,000, 2,000, 5,000,
10,000). Additional sequencing coverage would be expected to result in
more robust distance values since there would be more information
represented in the data. Indeed, the added sequencing depth showed
higher accuracy values at lower effect sizes for the combinations of
normalization methods and distance calculations (Figure S3). Increased
sequencing coverage also resulted in improved clustering accuracy for
lower effect sizes when the library size minimum quantile was decreased
(Figure S4). I will revisit the choice of the library size minimum
quantile below.

@@

A more robust analysis would have used rarefaction since it would have
averaged across a large number of random subsamplings (e.g., 100 or
1,000). By using a large number of subsamplings the likelihood of
incorporating all of the OTUs would have increased. Rather than being
guilty of ``omission of available valid data'' as claimed in WNWN, with
a sufficient number of subsamplings, traditional rarefaction uses all of
the available data. To fairly compare subsampling, as employed in WNWN,
and rarefaction, I removed the 15\% of samples with the lowest number of
sequences and compared the clustering accuracies from a single
subsampling to rarefaction with 100 randomizations. This analysis
revealed two benefits of rarefaction. First, the median distances
generated by rarefaction was always at least as large as those from a
single subsample (Figure 1). The difference was most pronounced for
smaller average library sizes and at smaller effect sizes. The
Unweighted UniFrac distances were most impacted by the use of
rarefaction over subsampling. Second, the interquartile ranges in
clustering accuracy by rarefaction were generally smaller than those by
subsampling and showed similar trends to the difference in the median
distances (Figure 1). Because rarefaction incorporates more of the data
and generally performed better than subsampling, the remainder of this
analysis will report results using rarefaction rather than by
subsampling, except when noted.

@@

Using the \texttt{nr-s1} Conda environment, I measured the richness and
Shannon diversity using the normalized OTU counts. Richness was measured
by counting the number of OTUs for each sample across the simulations
and the Shannon diversity using their relative abundance using the
commonly used formula (30). For the rarefaction data, each of the 80
samples in the 100 subsampled datasets were used to calculate the
richness and diversity. The mean across the 100 subsamplings was used as
the value for the rarefaction data across each of the 80 samples. To
assess differences between the two treatment groups, the richness and
diversity values were compared using the two-sample Wilcoxon
non-parametric test with the \texttt{wilcox.test} from the
\texttt{stats} base R package. Similar to the analysis using
\texttt{adonis2}, the fraction of significant tests was the fraction of
the 100 replicate simulations that had a p-value less than or equal to
0.05. The Type I error rate for a condition was the fraction of
significant tests when the effect size was 1.00. The power was the
fraction of significant tests at other effect sizes.

@@

\textbf{Figure S1. Re-running the R markdown files provided in Protocol
S1 of WNWN qualitatively reproduced Figure 4 from WNWN.}
